Commutative Character Sheaves

### 2015-09-28 followup.tex
Radical restructuring aimed at proving CCS(G) = CCS(Gab)

\section{Definitions and Recollections}

@@

Write $m : G \times G \to G$ for the multiplication morphism.
Let $\theta : G\times G \to G\times G$ be given by $\theta(g,h) = (h,g)$
and define $m\op : G \times G \to G$ by $m\op = m \circ \theta$.
For any local system $\gcs{L}$ on $\bG \ceq G \times_{\Spec{\Fq}} \Spec{\bFq}$ there is an isomorphism
$\vartheta : \LxL \to {\bar \theta}^*(\LxL)$ given on stalks by the canonical map
$\gcs{L}_{g} \otimes \gcs{L}_{h} \to \gcs{L}_{h} \otimes \gcs{L}_{g}$.

@@

\begin{definition}\label{def:CCS}
A rank-$1$ character sheaf on $G$ is \emph{commutative} if there is an isomorphism ${\bar m}^* \gcs{L} \to {\bar m}\op^* \gcs{L}$
such that the following diagram commutes,
  \[
  \begin{tikzcd}[row sep=30]
   \bm^*\gcs{L} \arrow{d} \arrow{r}{\mu} & \LxL \arrow{d}{\vartheta}\\
   \bm\op^*\gcs{L} \arrow{r}{{\bar \theta}^*\mu} &  {\bar \theta}^*(\LxL)
  \end{tikzcd}
  \]
 Write $\CCS(G)$ for the full subcategory of $\CS(G)$ consisting of commutative character sheaves.
 \end{definition}

\begin{lemma}
If the isomorphism ${\bar m}^* \gcs{L} \to {\bar m}\op^* \gcs{L}$ exists, then it is unique.
\end{lemma}
\begin{proof}
\end{proof}

\begin{proposition}
Every commutative character sheaf on $G$ is equivariant for the conjugation action of $G$ on $G$. Conversely, if a rank-$1$ character sheaf on $G$ is equivariant for conjugation, then it is commutative. 
\end{proposition}

@@

\subsection{Dominant morphisms and local systems}

\begin{lemma}\label{lem:dominant}
Suppose $f: X \to Y$ is dominant.
If $\cs{L}\in \Loc(Y)$ and $f^*\cs{L} \iso (\EE)_X$ then $\cs{L} \iso (\EE)_Y$.
\end{lemma}

\begin{proof}
The group homomorphism $\pi_1(f): \pi_1(X, {\bar x}) \to \pi_1(Y, {\bar y})$ is surjective, for all geometric points with $f({\bar x}) = {\bar y}$ ... 
\todo{Under construction}
\end{proof}

\subsection{Derived subgroup}

Let $G$ be a smooth group scheme.
Let $j: G\der \hookrightarrow G$ denote the closed subgroup generated by the commutator morphism $c: G\times G \to G$ given by $c(x,y) = xyx^{-1}y^{-1}$; see \cite{SGA3}*{{VI}$_B$, 7.2}.

\[
\begin{tikzcd}
{} & G\times G \arrow{d}{c} &&&\\
1 \arrow{r} & G\der \arrow{r}{j} & G \arrow{r}{q} & G\ab \arrow{r} & 1
\end{tikzcd}
\]

@@

\begin{remark}
Lemma~\ref{lem:Lder} provides another indication of the correctness of Definition~\ref{def:CCS} 
if one is looking for the geometrization of characters of smooth group schemes over finite fields: commutative character sheaves are trivial on $G\der$, rank-$1$ character sheaves are not.
In the next section we provide examples to show that $\CCS(G)$ and $\CS(G)$ are not isomorphic categories, even when $G$ is a connected reductive algebraic group. 
\end{remark}

@@

\section{Descent for commutative character sheaves}

\subsection{Quotients}

\subsection{$H$-equivariance}

An $H$-equivariant rank-$1$ character sheaf on $G$ is a rank-$1$ character sheaf $\cs{L} = (\gcs{L},\phi,\mu)$ on $G$ together with an isomorphism  $\nu : {\bar m}_H^*\gcs{L} \to {\bar p}_H^*\gcs{L}$ such that the following diagrams commute, expressing compatibility with $\phi$ and $\mu$:
\begin{enumerate}
\item
\[
\begin{tikzcd}
\ &  m_0^*  m^*\, \mathcal{L} \arrow{r}{m_0^*\nu} & m_0^* p^*\, \mathcal{L} & \\
m_1^* m^*\, \mathcal{L} \arrow{dr}[swap]{m_1^*\nu} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{p_2^*\nu} p_2^* m^*\, \mathcal{L} \\
& m_1^* p^*\, \mathcal{L} \arrow[equal]{r} & p_2^* p^*\, \mathcal{L} & 
\end{tikzcd}
\]
where $m_0(a,b,x) = (a,m(b,x))$ and $m_1(a,b,x) = (m(ab),x)$ and $p_2(a,b,x) = (b,x)$;
\item
\end{enumerate}
Write $\CCS_H(G)$ for the category of $H$-equivariant commutative character sheaves on $G$.

\begin{proposition}\label{prop:Gder-equivariance}
$\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence.
\end{proposition}

\begin{proof}
%Let $j : C \hookrightarrow G$ be the kernel of $q : G \to G\ab$ and let $m : G\times G \to G$ be multiplication for $G$. 
Let $n : G\der \times G \to G$ be the restriction of $m: G\times G \to G$.
Then $n = m \circ (j \times \id)$, so
\begin{align*}
n^* \cs{L} 
&= (m \circ (j \times \id))^* \cs{L}\\
&= (j \times \id))^* m^*\cs{L}\\
&\iso (j \times \id))^* (\cs{L} \boxtimes \cs{L})\\
&= (j^*\cs{L}) \boxtimes \cs{L}\\
&= (\cs{L}\vert_C) \boxtimes \cs{L}\\
&= (\EE)_C \boxtimes \cs{L}\\
&= p^* \cs{L}.
\end{align*}
Thus, every commutative character sheaf $\cs{L}$ on $G$ is $G\der$-equivariant.
It now follows from Lemma~\ref{lem:Lder} that the forgetful functor $\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence. 
\end{proof}

We now suppose that $H$ is normal in $G$ and that the fppf-quotient $G \to G'\ceq H\backslash G$ exists.
Then $q : G \to G'$ is a regular epimorphism with kernel pair $(n,p)$ in the category of schemes.
In particular, $G \times_{G'} G \iso H\times_S G'$, so $G \to G'$ is a $G$-torsor in the fppf topology.

In this case we may factor the functor $\CCS(G') \to \CCS(G)$  . . . 
\[
\begin{tikzcd}
\CCS(G) &&\arrow{ll}[swap]{q^*} 
\CCS(G') \arrow{dl}{\text{`comparison'}}\\
& \arrow{ul}{\text{forget}} \CCS_{H}(G) & 
\end{tikzcd}
\]

\begin{lemma}\label{lem:descent}
If the quotient $q : G \to G'$ is a $G$-torsor in the etale topology, then the comparison functor $\CCS(G') \to \CCS_{H}(G)$ is an equivalence.
\end{lemma}

\begin{proof}
By descent theory, the functor $\Loc(G') \to \Loc_H(G)$ induced by $q: G \to G'$ is an equivalence of categories.
\todo{Need to study the extra structure coming with character sheaves.}
\end{proof}

By definition, the abelianization $G\ab$ of $G$ is the fppf-quotient $q : G \to G\ab$ of $G$ by $G\der$. 
This exists in schemes and is a $G\der$-torsor in the etale topology since  . . . \todo{Either true or it becomes a condition on $G$. I hope smoothness does the trick. I've written to Masoud for help on this.}
%\todo{Proof needed}
%  is the coequalizer of $n : G\der \times G \to G$ with projection $p: G\der \times G \to G$. 
\[
\begin{tikzcd}
G\der\times G
 \arrow{r}[shift left]{n}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G\ab
\end{tikzcd}
\]

\begin{proposition}\label{prop:Gab-descent}
Let $G$ be a smooth group scheme over $\Fq$ 
 and let $q : G \to G\ab$ be its abelianization. 
 Then  
\[
\CCS(G\ab) \longrightarrow \CCS_{G\der}(G)
\]
is an equivalence of categories.
\end{proposition}

\begin{proof}
The quotient $G \to G\ab$ is a $G\der$-torsor in the etale topology,  so the result follows from Lemma~\ref{lem:descent}.
\end{proof}

\begin{theorem}\label{thm:Gab}
 Let $G$ be a smooth group scheme over $\Fq$ 
 and let $q : G \to G\ab$ be its abelianization. 
 Then  
\[
q^*:  \CCS(G\ab) \longrightarrow \CCS(G)
\]
is an equivalence of rigid monoidal categories.
\end{theorem}
\begin{proof}
Combine Propositions~\ref{prop:Gder-equivariance} and \ref{prop:Gab-descent}.
\end{proof}

 \begin{corollary}
The following diagram of isomorphisms commutes.
\[
\begin{tikzcd}
\CCSiso{G} \arrow{r} & G(\Fq)^*\\
\arrow{u}{q^*} \CCSiso{G\ab} \arrow{r} & \arrow{u} G\ab(\Fq)^* 
\end{tikzcd}
\]
 \end{corollary}

 \begin{proof}
 By Theorem \ref{thm:trfrobiso}, $\CCSiso{G}\to G(\Fq)^*$ and $\CCSiso{G\ab}\to G\ab(\Fq)^*$ are isomorphisms.
By Theorem~\ref{thm:Gab}, $\CCSiso{G} \iso \CCSiso{G\ab}$.
Since $G\der$ is simply connected, \todo{Proof needed} $G\ab(\Fq) = G(\Fq)/G\der(\Fq)$.
Thus, all characters of $G(\Fq)$ factor through
 $G\ab(\Fq)$. . . . 
 The isomorphisms are compatible because  . . . 
 \end{proof}

\begin{remark} 
It is not true that the categories $\CS(G\ab)$ and $\CS(G)$ are equivalent, even when $G$ is a connected reductive algebraic group, as we will see in the next subsection.
\end{remark}

\subsection{Stacky abelianization, revisited}

>>>>

\section{Main Definition and Result}

@@

\subsection{Main definition}

Let $\theta : G\times G \to G\times G$ be given by $\theta(g,h) = (h,g)$.
Write $m : G \times G \to G$ for the multiplication morphism
and define $m\op : G \times G \to G$ by $m\op = m \circ \theta$.
For any local system $\gcs{L}$ on $\bG \ceq G \times_{\Spec{\Fq}} \Spec{\bFq}$ there is an isomorphism
$\vartheta : \LxL \to {\bar \theta}^*(\LxL)$ given on stalks by the canonical map
$\gcs{L}_{g} \otimes \gcs{L}_{h} \to \gcs{L}_{h} \otimes \gcs{L}_{g}$.

@@

\begin{definition}\label{def:CCS}
A rank-$1$ character sheaf on $G$ is a\emph{commutative character sheaf} if there is an isomorphism ${\bar m}^* \gcs{L} \to {\bar m}\op^* \gcs{L}$ such that the following diagram commutes,
  \[
  \begin{tikzcd}[row sep=30]
   \bm^*\gcs{L} \arrow{d} \arrow{r}{\mu} & \LxL \arrow{d}{\vartheta}\\
   \bm\op^*\gcs{L} \arrow{r}{{\bar \theta}^*\mu} &  {\bar \theta}^*(\LxL).
  \end{tikzcd}
  \]
 Write $\CCS(G)$ for the full subcategory of $\CS(G)$ consisting of commutative character sheaves.
 \end{definition}

\begin{lemma}
If the isomorphism ${\bar m}^* \gcs{L} \to {\bar m}\op^* \gcs{L}$ exists and satisfies the condition of Definition~\ref{def:CCS}, then it is unique.
\end{lemma}
\begin{proof}
\end{proof}

\subsection{Commutative group schemes}

\begin{lemma}\label{lem:commutative}
If $G$ is commutative then the categories $\CCS(G)$ and $\CS(G)$ are equivalent.
\todo{I'm confused. There are extra isomorphisms in the former category, but I think they're equivalent for indirect reasons.}
\end{lemma}

Let $G$ be a smooth group scheme.
Let $G\der \hookrightarrow G$ denote the closed subgroup generated by the smooth morphism $G\times G \to G$ given by $(x,y)\mapsto xyx^{-1}y^{-1}$. Then $G \times G \to G\der$ is a dominant morphism; see \cite{SGA3}*{{VI}$_B$, 7.2}.
\[
\begin{tikzcd}
{} & G\times G \arrow{d}{} &&&\\
1 \arrow{r} & G\der \arrow{r}{} & G \arrow{r}{q} & G\ab \arrow{r} & 1
\end{tikzcd}
\]
Then $G$ is commutative if and only if $G\der$ is

While the subgroup $G\der$ is smooth, it need not be connected. The restriction of $G\times G\to G\der$ to identity components gives a dominant morphism $G^0\times G^0 \to G^0\der$ by [SGA6B, Ex. 7, Prop 7.8]

Let $q: G \to G\ab$ be the fppf-quotient  . . . 
We suppose $G\ab$ exists in group schemes over $\Fq$. \todo{This is a condition on $G$. Does it follow from smoothness? I've written to Masoud for help on this.}
If $G$ is of finite type (and so, an algebraic group) then $G\ab$ exists in group schemes by \cite{GM}*{Thm 4.39}: $G$ smooth implies $G$ flat; $G\der$ is also smooth (!) so also flat; $\dim \Spec{\Fq} \leq 1$.

 \subsection{Statement of the main result}

As in [CS], every morphism of group schemes $G \to G'$ induces . .

Another hypothesis: suppose the smooth morphism $G\times G\to G\der$ is of relative dimension on each connected component of $G\der$.

\begin{theorem}\label{thm:Gab}
Pull-back along the abelianization $q : G \to G\ab$
\[
q^* : \CCS(G\ab) \to \CCS(G)
\]
is an equivalence of categories.
\end{theorem}

Theorem~\ref{thm:Gab} will be proved by factoring $\CCS(G\ab) \to \CCS(G\ab)$ according to the following diagram, which will be explained below, 
and by showing that each functor is an equivalence.
\[
\begin{tikzcd}
 \CCS(G)  &&& \arrow{lll}[swap]{q^*}  \CCS(G\ab) \arrow{d}{\text{Equivariance}} \\
\CS_G(G)  \arrow{u}{\text{Equivariance}} &&& \CS_{G\ab}(G\ab) \arrow{d}{\text{Commutative group}}  \\
\CS_{G\der}(G) \arrow{u}{\text{Change of Action}} &&& \arrow{lll}{\text{Descent along torsor}}  \CS(G\ab)  \\
\end{tikzcd}
\]

\begin{remark}
Theorem~\ref{thm:Gab} provides an indication that $\CCS(G)$ is the correct category in which to geometrize characters of $G(\Fq)$. 
By contrast, $\CS(G)$ and $\CS(G\ab)$ are generally not isomorphic categories, not even when $G$ is a connected reductive algebraic group and $\Hh^1(\Fq,G\der)=0$, as we will see in Section~\ref{sec:stacky}.
\end{remark}

\subsection{Equivariance}

We wish to consider the category $\CS_G(G)$ of $G$-equivariant linear character sheaves on $G$. To that end, in this section we introduce that notion in its general context.

Let $G$ and $H$ be group schemes; let $n : H \times G\to G$ be a group action and write $p : H\times G \to G$ for projection. 
An $H$-equivariant linear character sheaf on $G$ is a linear character sheaf $\cs{L} = (\gcs{L},\phi,\mu)$ on $G$ together with an isomorphism  $\nu : {\bar n}^*\gcs{L} \to {\bar p}^*\gcs{L}$ such that the following diagrams commute, expressing compatibility with $\phi$ and $\mu$:
\begin{enumerate}
\item
\[
\begin{tikzcd}
\ &  n_0^*  n^*\, \mathcal{L} \arrow{r}{n_0^*\nu} & n_0^* p^*\, \mathcal{L} & \\
n_1^* n^*\, \mathcal{L} \arrow{dr}[swap]{n_1^*\nu} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{p_2^*\nu} p_2^* n^*\, \mathcal{L} \\
& n_1^* p^*\, \mathcal{L} \arrow[equal]{r} & p_2^* p^*\, \mathcal{L} & 
\end{tikzcd}
\]
where $n_0(a,b,x) = (a,n(b,x))$ and $n_1(a,b,x) = (n(ab),x)$ and $p_2(a,b,x) = (b,x)$;
\item
\end{enumerate}
Write $\CCS_H(G)$ for the category of $H$-equivariant commutative character sheaves on $G$.

\begin{lemma}\label{lem:equivariant}
$\CCS(G)$ and $\CS_G(G)$ are equivalent.
\end{lemma}
%\begin{proposition}
%Every commutative character sheaf on $G$ is equivariant for the conjugation action of $G$ on $G$. Conversely, if a rank-$1$ character sheaf on $G$ is equivariant for conjugation, then it is commutative. 
%\end{proposition}

@@

\begin{remark}
Lemma~\ref{lem:equivariant} shows that the `canonical functor' $\CCS(G) \to \CS(G)$ is a forgetful functor and also provides an alternate characterization of commutative character sheaves as linear (rank-$1$), multiplicative ($\mu$) and conjugation-equivariant ($\nu$) Weil local systems ($\phi$) on $\bG$.
\end{remark}

We now suppose that $H$ is normal in $G$ and that the fppf-quotient $G \to G'\ceq H\backslash G$ exists in group schemes over $\Fq$.
Then $q : G \to G'$ is a regular epimorphism with kernel pair $(n,p)$ in the category of schemes.
\[
\begin{tikzcd}
H\times G
 \arrow{r}[shift left]{n}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G'
\end{tikzcd}
\]
%In particular, $G \times_{G'} G \iso H\times_S G'$, so $G \to G'$ is a $G$-torsor in the fppf topology.
In this case we may factor the functor $\CCS(G') \to \CCS(G)$  . . . 
\[
\begin{tikzcd}
\CCS(G) &&\arrow{ll}[swap]{q^*} 
\CCS(G') \arrow{dl}{\text{`comparison'}}\\
& \arrow{ul}{\text{forget}} \CCS_{H}(G) & 
\end{tikzcd}
\]

\begin{lemma}\label{lem:descent}
The comparison functor $\CCS(G') \to \CCS_{H}(G)$ is an equivalence.
\end{lemma}

\begin{proof}
By descent theory, the functor $\Loc(G') \to \Loc_H(G)$ induced by $q: G \to G'$ is an equivalence of categories. Expand . . . 
\todo{Need to study the extra structure coming with character sheaves.}

%Let $j : C \hookrightarrow G$ be the kernel of $q : G \to G\ab$ and let $m : G\times G \to G$ be multiplication for $G$. 
Let $n : G\der \times G \to G$ be the restriction of $m: G\times G \to G$.
Then $n = m \circ (j \times \id)$, so
\begin{align*}
n^* \cs{L} 
&= (m \circ (j \times \id))^* \cs{L}\\
&= (j \times \id))^* m^*\cs{L}\\
&\iso (j \times \id))^* (\cs{L} \boxtimes \cs{L})\\
&= (j^*\cs{L}) \boxtimes \cs{L}\\
&= (\cs{L}\vert_C) \boxtimes \cs{L}\\
&= (\EE)_C \boxtimes \cs{L}\\
&= p^* \cs{L}.
\end{align*}
Thus, every commutative character sheaf $\cs{L}$ on $G$ is $G\der$-equivariant.
It now follows from Lemma~\ref{lem:Lder} that the forgetful functor $\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence. 
\end{proof}

\subsection{Change of action}

\[
\begin{tikzcd}
\arrow{dd}[swap]{f} G_1 \times G \arrow{dr}[swap]{a_1}{p} & \\
& G \\
G_2 \times G \arrow{ur}{a_2}[swap]{p} & 
\end{tikzcd}
\]

\begin{lemma}\label{lem:action}
The canonical functor
\[
\CS_{G_2}(G) \to \CS_{G_1}(G)
\]
is an equivalence if $f$ is smooth of relative dimension on connected components.
\end{lemma}
\begin{proof}
\end{proof}

\subsection{Proof of the main result}

\begin{proof}[Proof of Theorem~\ref{thm:Gab}]
First we must see that $q^*$ factors according to . . . 
\[
\begin{tikzcd}
 \CCS(G)  &&& \arrow{lll}[swap]{q^*}  \CCS(G\ab) \arrow{d}{\text{Lem~\ref{lem:equivariant}}} \\
\CS_G(G)  \arrow{u}{\text{Lem~\ref{lem:equivariant}}} &&& \CS_{G\ab}(G\ab) \arrow{d}{\text{Lem~\ref{lem:commutative}}}  \\
\CS_{G\der}(G) \arrow{u}{\text{Lem~\ref{lem:action}}} &&& \arrow{lll}{\text{Lem~\ref{lem:descent}}}  \CS(G\ab)  \\
\end{tikzcd}
\]
\end{proof}

@@

\section{Geometrization of linear characters}

 \begin{corollary}
 If $\Hh^1(\Fq,G\der)=0$ then
the following diagram of isomorphisms commutes.
\[
\begin{tikzcd}
\CCSiso{G} \arrow{r} & G(\Fq)^*\\
\arrow{u}{q^*} \CCSiso{G\ab} \arrow{r} & \arrow{u} G\ab(\Fq)^* 
\end{tikzcd}
\]
 \end{corollary}

 \begin{proof}
 By Theorem \ref{thm:trfrobiso}, $\CCSiso{G}\to G(\Fq)^*$ and $\CCSiso{G\ab}\to G\ab(\Fq)^*$ are isomorphisms.
By Theorem~\ref{thm:Gab}, $\CCSiso{G} \iso \CCSiso{G\ab}$.
Since $\Hh^1(\Fq,G\der)=0$, we have  $G\ab(\Fq) = G(\Fq)/G\der(\Fq)$.
Thus, all characters of $G(\Fq)$ factor through
 $G\ab(\Fq)$. . . . 
 The isomorphisms are compatible because  . . . 
 \end{proof}

@@

\section{Stacky abelianization, revisited}\label{sec:stacky}
